\chapter{Each Tool in Brief}
\label{ch:each-tool-in-brief}

Now comes something remarkable. Having listed his tools in Chapter~2, Mahākaccāyana does something that no modern textbook would think to do: he compresses each one into a two-line verse.\footnote{This section is the \textit{Niddesavāro} --- literally ``the description section'' or ``the exposition section.'' The Pali opens with a single line: \textit{``Tattha saṅkhepato netti kittitā''} --- ``Herein the Guide has been proclaimed in brief.'' Then come twenty-six numbered verses covering all sixteen modes, five methods, and the twelve building blocks of a sutta. The word \textit{niddesa} (from \textit{ni} + \textit{disati}, to point out) means a description or exposition --- distinct from the bare enumeration (\textit{uddesa}) of Chapter~2. The Netti's own three-layer structure --- uddesa, niddesa, paṭiniddesa --- mirrors the pedagogical method it teaches: first list, then summarize in verse, then explain in full.}

Why? Because this is an oral tradition. These aren't verses to be admired on the page --- they're \textit{mnemonics}. Two thousand years ago, a student learning the Netti would memorize these twenty-six couplets and carry them in their head the way a medical student carries the cranial nerves: as a compressed index that unpacks into a vast body of knowledge when you need it.

Think of it this way. Chapter~2 was the syllabus. This chapter is the set of flash cards.\footnote{The three-part structure of \textit{saṅgahavāro}--\textit{uddesavāro}--\textit{niddesavāro} (overview--enumeration--description) is itself a teaching technique the Netti identifies and uses. First you get the bird's-eye view, then the list, then the compressed definitions. Only after all three does the detailed analysis begin (the \textit{paṭiniddesavāro}, Chapters~4--16). It's like a textbook that gives you the table of contents, then the chapter summaries, then the full chapters.}

\ornament

\section*{The Sixteen Modes in Verse}

Each mode gets a single couplet --- a definition so compressed it reads like a telegram. Here they are, one by one, with each verse followed by a brief unpacking of what it's saying.

\bigskip

\subsection*{1. Teaching}

\begin{versebox}
Appeal and danger, escape and fruit and means,\\
plus the Buddha's direct command to practitioners ---\\
that's the Teaching Mode.
\end{versebox}

Six ingredients. Whenever the Buddha teaches about anything --- sense pleasure, a mental state, a practice --- he runs through the same six-step program: here's what's attractive about it, here's what's dangerous about it, here's how to get free of it, here's the result of getting free, here's the method, and now --- do it.\footnote{Verse 1. Pali: \textit{``Assādādīnavatā, nissaraṇampi ca phalaṃ upāyo ca; / Āṇattī ca bhagavato, yogīnaṃ desanāhāro.''} The six elements: \textit{assāda} (gratification, appeal), \textit{ādīnava} (danger, drawback), \textit{nissaraṇa} (escape, way out), \textit{phala} (fruit, result), \textit{upāya} (means, method), and \textit{āṇatti} (injunction, command). The last element is striking: the Buddha doesn't just describe --- he \textit{instructs}. This is not philosophy. It's a training program.}

That last element --- the direct command --- is key. The Buddha isn't a philosopher musing about the nature of things. He's a teacher giving instructions. ``Develop this.'' ``Abandon that.'' ``Practice this way.'' The Teaching Mode captures the structure of those instructions.\footnote{The word \textit{āṇatti} (injunction, command) appears frequently in the Vinaya (the monastic code) but is less common in sutta contexts. Its use here underlines the Netti's view that the suttas are not just descriptive but \textit{prescriptive}: they don't merely analyze experience, they direct action. Compare the modern Buddhist tendency to see meditation instructions as ``invitations'' --- the Netti sees them as commands from a teacher who knows what works.}

\subsection*{2. Investigation}

\begin{versebox}
What question was asked? What answer was given?\\
And what verse sums it up?\\
Investigate the sutta --- that's this mode.
\end{versebox}

Three components: the question, the answer, and the summary verse. Most suttas follow this pattern --- someone asks the Buddha something, the Buddha responds, and a verse at the end crystallizes the teaching.\footnote{Verse 2. Pali: \textit{``Yaṃ pucchitañca vissajjitañca, suttassa yā ca anugīti; / Suttassa yo pavicayo, hāro vicayoti niddiṭṭho.''} Three elements: \textit{pucchita} (what was asked), \textit{vissajjita} (what was answered), and \textit{anugīti} (the follow-up verse, literally ``sung along with'' --- from \textit{anu} + \textit{gīti}, singing). The term \textit{pavicaya} (investigation, analysis) is from the same root as \textit{dhammavicaya}, the investigation-of-states factor of awakening. Investigating a sutta has the same quality as investigating phenomena in meditation --- you're looking for structure, not just content.}

\subsection*{3. Fitness}

\begin{versebox}
What terrain do all the modes work on?\\
What's their range?\\
Testing fit from unfit --- that's the Fitness Mode.
\end{versebox}

This is the meta-mode. It doesn't analyze a sutta directly --- it checks whether your \textit{analysis} is any good. Does your reading actually fit? Is the mode you're using appropriate for this passage? Are you applying it within its proper range?\footnote{Verse 3. Pali: \textit{``Sabbesaṃ hārānaṃ, yā bhūmī yo ca gocaro tesaṃ; / Yuttāyuttaparikkhā, hāro yuttīti niddiṭṭho.''} Key terms: \textit{bhūmi} (ground, terrain --- what kind of text each mode applies to), \textit{gocara} (range, pasture --- how far each mode extends), and \textit{yuttāyutta} (fit and unfit). The metaphor is pastoral: each mode has its ``grazing ground'' and you need to know the boundaries. An interpretation that wanders outside the mode's range is ``unfit'' --- it may be clever, but it doesn't belong here.}

Every interpretive system needs a quality check, and this is the Netti's. You can't just apply any mode anywhere and call the result valid. The Fitness Mode ensures disciplined reading.\footnote{Ñāṇamoli notes that \textit{yutti} (fitness) is the Netti's principle of interpretive validity. A reading must be \textit{yutta} (fitting) not just in isolation but in relation to the entire body of teaching. This anticipates by two millennia the hermeneutic principle that any interpretation of a part must be consistent with the whole.}

\subsection*{4. Basis}

\begin{versebox}
The Victor teaches a quality ---\\
and this mode finds what it stands on.\\
Trace it all the way down: that's the Basis Mode.
\end{versebox}

When the Buddha teaches about, say, right mindfulness, what does mindfulness rest on? What supports it? What is its foundation? This mode traces any teaching down to its roots.\footnote{Verse 4. Pali: \textit{``Dhammaṃ deseti jino, tassa ca dhammassa yaṃ padaṭṭhānaṃ; / Iti yāva sabbadhammā, eso hāro padaṭṭhāno.''} \textit{Jino} (the Victor, the Conqueror) is a common epithet for the Buddha. \textit{Padaṭṭhāna} literally means ``foot-standing-place'' (\textit{pada} + \textit{ṭhāna}) --- the ground on which something stands. The phrase \textit{yāva sabbadhammā} (``all the way to all dhammas'') indicates that this mode is recursive: you trace the basis of a quality, then the basis of \textit{that} basis, and so on until you've mapped the entire foundation.}

The verse says ``all the way down'' --- \textit{yāva sabbadhammā}. This mode is recursive. You don't just find one foundation; you keep tracing until you hit bedrock. What does compassion stand on? Perhaps non-hatred. What does non-hatred stand on? And so on.\footnote{This recursive quality makes the Basis Mode one of the most powerful analytical tools in the Netti. It effectively maps the causal hierarchy of the path. In Abhidhamma terms, it traces the \textit{upanissaya-paccaya} (decisive-support condition) chain. In modern terms, it's a root-cause analysis.}

\subsection*{5. Characteristic}

\begin{versebox}
When one quality is stated, then every quality\\
that shares its mark is stated too ---\\
that's the Characteristic Mode.
\end{versebox}

This is a principle of extraordinary power. When the Buddha describes the impermanence of one thing, he has implicitly described the impermanence of \textit{everything that is impermanent}. The teaching about one thing radiates outward to everything that shares the same defining mark.\footnote{Verse 5. Pali: \textit{``Vuttamhi ekadhamme, ye dhammā ekalakkhaṇā keci; / Vuttā bhavanti sabbe, so hāro lakkhaṇo nāma.''} \textit{Ekalakkhaṇā} means ``having one characteristic'' or ``sharing a single mark.'' The logical structure here is essentially universal generalization: when a property is stated of one instance, it holds for all instances that share the relevant characteristic. This is remarkably close to modern predicate logic's universal quantification.}

This explains why the Buddha can teach about ``the impermanence of form'' and have that teaching simultaneously cover the impermanence of feeling, perception, formations, and consciousness. He doesn't always need to spell out all five aggregates --- stating one that shares the characteristic covers them all.\footnote{This principle also explains the famous ``formula'' style of the suttas, where the Buddha gives a teaching about one item and then says ``and so too for the remaining ones'' (\textit{iti evaṃ sabbattha}). What looks like laziness or repetitiveness is actually the Characteristic Mode in action: the Canon assumes the reader will generalize from the characteristic to all items that share it.}

\subsection*{6. The Fourfold Analysis}

\begin{versebox}
Grammar, intention, phrasing and its source,\\
and the thread from before to after ---\\
four arrays: the Fourfold Analysis Mode.
\end{versebox}

Four lenses, applied simultaneously to any passage: What do the words literally mean? What's the speaker's underlying intention? What's the context and background? And how does this passage connect to what comes before and after it?\footnote{Verse 6. Pali: \textit{``Neruttamadhippāyo, byañjanamatha desanānidānañca; / Pubbāparānusandhī, eso hāro catubyūho.''} The four elements: (1) \textit{nerutti} (grammatical analysis, etymology --- from \textit{nir} + \textit{vac}, to speak forth), (2) \textit{adhippāya} (intention, purpose), (3) \textit{byañjana} + \textit{desanānidāna} (phrasing and the teaching's origin/occasion), and (4) \textit{pubbāparānusandhi} (the connection of what precedes to what follows --- literally ``before-after-linking''). This is essentially the ancient equivalent of modern discourse analysis: linguistics, pragmatics, context, and textual coherence in one integrated method.}

The term ``fourfold array'' (\textit{catubyūha}) comes from military language --- it's a four-part battle formation. You deploy all four lenses at once, like approaching a text from four directions simultaneously.\footnote{\textit{Byūha} literally means ``arrangement, array, formation.'' In military contexts it refers to a strategic deployment of forces. The metaphor suggests that understanding a sutta requires a coordinated multi-pronged approach, not just a single line of attack. Ñāṇamoli translated the term as ``four-fold array'' and noted its military resonance.}

\subsection*{7. Turning Back}

\begin{versebox}
From one basis, seek the remaining basis;\\
then turn to the opposite ---\\
that's the Turning Back Mode.
\end{versebox}

Two moves: trace the foundation (using the Basis Mode), then \textit{flip}. Whatever you found on the dark side, turn to its bright counterpart. If you traced a teaching down to craving, now turn to the opposite --- non-craving.\footnote{Verse 7. Pali: \textit{``Ekamhi padaṭṭhāne, pariyesati sesakaṃ padaṭṭhānaṃ; / Āvaṭṭati paṭipakkhe, āvaṭṭo nāma so hāro.''} \textit{Sesaka} means ``remaining'' --- after finding one basis, you search for the rest. Then \textit{āvaṭṭati paṭipakkhe} --- ``it turns to the opposite.'' The image is of a wheel turning: you follow the chain one way, then revolve it to face the other direction. This is how the Netti connects the diagnosis (unwholesome roots) to the cure (wholesome roots) within any single teaching.}

This mode is the bridge between problem and solution. The Buddha's teaching always has two sides --- what to abandon and what to develop --- and the Turning Back Mode makes you see both from any starting point.

\subsection*{8. Sorting}

\begin{versebox}
The quality, its basis, and its ground ---\\
this mode sorts shared from unshared.\\
That's the Sorting Mode.
\end{versebox}

Not everything in a sutta applies to everyone equally. Some teachings are universal (``suffering exists''), some are specific to certain stages of practice (``enter the first jhāna''). The Sorting Mode divides teachings into shared and unshared applications.\footnote{Verse 8. Pali: \textit{``Dhammañca padaṭṭhānaṃ, bhūmiñca vibhajjate ayaṃ hāro; / Sādhāraṇe asādhāraṇe ca neyyo vibhattīti.''} Three things are sorted: the \textit{dhamma} (quality, teaching), the \textit{padaṭṭhāna} (basis), and the \textit{bhūmi} (ground, level). Each is classified as \textit{sādhāraṇa} (shared, common, universal) or \textit{asādhāraṇa} (unshared, specific, particular). This prevents the error of universalizing a specific instruction or, conversely, limiting a universal teaching to one context.}

\subsection*{9. Reversal}

\begin{versebox}
Wholesome and unwholesome qualities ---\\
described, developed, abandoned ---\\
this mode flips to the opposite. That's Reversal.
\end{versebox}

When the Buddha says ``develop loving-kindness,'' what is being abandoned? Ill-will. When he says ``abandon ill-will,'' what is being developed? Loving-kindness. Every positive instruction has a negative shadow, and the Reversal Mode makes both explicit.\footnote{Verse 9. Pali: \textit{``Kusalākusale dhamme, niddiṭṭhe bhāvite pahīne ca; / Parivattati paṭipakkhe, hāro parivattano nāma.''} Three things are done with wholesome and unwholesome qualities: they are \textit{niddiṭṭha} (described, pointed out), \textit{bhāvita} (developed, cultivated), and \textit{pahīna} (abandoned, given up). Then the mode \textit{parivattati paṭipakkhe} --- ``turns to the opposite.'' \textit{Parivattana} means ``turning around, reversal'' --- the same root as in \textit{dhammacakkappavattana}, the ``turning of the wheel of Dhamma.''}

\subsection*{10. Synonyms}

\begin{versebox}
Many synonyms for a single thing\\
are stated across the suttas ---\\
the sutta-expert knows them. That's the Synonym Mode.
\end{versebox}

The Buddha uses different words for the same concept in different suttas. ``Craving,'' ``delight,'' ``thirst,'' ``attachment'' --- these can all point to the same underlying process. The Synonym Mode maps these equivalences.\footnote{Verse 10. Pali: \textit{``Vevacanāni bahūni tu, sutte vuttāni ekadhammassa; / Yo jānāti suttavidū, vevacano nāma so hāro.''} \textit{Vevacana} means ``synonym, alternative expression'' (literally ``different word''). The key phrase is \textit{ekadhammassa} --- ``of one and the same thing.'' The verse says the \textit{suttavidū} (``one who knows the suttas'') recognizes these equivalences. This is essentially the Netti's term for cross-referencing: building a mental concordance across the entire Canon.}

This is one of the most practically useful modes. Without it, you read the suttas as thousands of separate talks. With it, you start seeing the same small number of core teachings expressed in an enormous variety of language.

\subsection*{11. Designation}

\begin{versebox}
The Blessed One teaches one reality\\
through various designations ---\\
know this method: the Designation Mode.
\end{versebox}

Similar to Synonyms, but operating at a higher level. Synonyms are different \textit{words} for the same thing. Designations are different \textit{frameworks} for the same reality. The five aggregates, the twelve sense bases, the eighteen elements --- these aren't competing descriptions. They're different maps of the same territory.\footnote{Verse 11. Pali: \textit{``Ekaṃ bhagavā dhammaṃ, paññattīhi vividhāhi deseti; / So ākāro ñeyyo, paññattī nāma hāroti.''} \textit{Paññatti} means ``designation, concept, name.'' The phrase \textit{ekaṃ dhammaṃ vividhāhi paññattīhi} --- ``one reality through various designations'' --- is a profound hermeneutic principle: the diversity of the Buddha's frameworks is \textit{pedagogical}, not ontological. He's not describing different realities; he's describing the same reality through different lenses for different students.}

\subsection*{12. Entry Point}

\begin{versebox}
Dependent origination, faculties,\\
aggregates, elements, sense bases ---\\
enter through these: the Entry Point Mode.
\end{versebox}

Any teaching can be mapped onto the standard doctrinal categories. Whatever the Buddha is talking about, you can always find a way in through dependent origination, the aggregates, the sense bases, the elements, or the faculties.\footnote{Verse 12. Pali: \textit{``Yo ca paṭiccuppādo, indriyakhandhā ca dhātu āyatanā; / Etehi otarati yo, otaraṇo nāma so hāro.''} The five entry frameworks: \textit{paṭiccasamuppāda} (dependent origination), \textit{indriya} (faculties), \textit{khandha} (aggregates), \textit{dhātu} (elements), and \textit{āyatana} (sense bases). The verb \textit{otarati} means ``to descend into, to enter'' --- the image is of diving below the surface of the sutta's specific language into the deeper doctrinal structures underneath.}

These five frameworks are the universal access points. No matter how unfamiliar or specialized a sutta seems, you can always find your way in by asking: how does this connect to dependent origination? Which aggregates are involved? Which sense bases?

\subsection*{13. Clearing}

\begin{versebox}
When the answer is given to a question\\
first posed in verse ---\\
checking pure from impure: that's the Clearing Mode.
\end{versebox}

Many suttas open with a question in verse and then give an answer in prose. Does the prose answer actually cover everything the verse asked? This mode checks. If the answer is ``pure'' (\textit{suddha}), it fully addresses the question. If ``impure'' (\textit{asuddha}), something was left out.\footnote{Verse 13. Pali: \textit{``Vissajjitamhi pañhe, gāthāyaṃ pucchitāyamārabbha; / Suddhāsuddhaparikkhā, hāro so sodhano nāma.''} \textit{Sodhana} means ``purification, clearing, verification'' --- from \textit{suddhi} (purity). The mode examines the relationship between a verse question (\textit{gāthāyaṃ pucchitāya}) and a prose answer (\textit{vissajjita}). This is essentially a completeness check: did the answer address all parts of the question?}

Think of it as debugging for sutta interpretation. If you think you understand what the Buddha was answering, this mode tests whether your understanding actually covers the question.

\subsection*{14. Determination}

\begin{versebox}
Qualities stated as sharing one nature,\\
and those stated as differing ---\\
sort them out: the Determination Mode.
\end{versebox}

When the Buddha describes two qualities, are they the same thing in different words, or genuinely different things? The Determination Mode forces you to decide --- and the answer matters for practice.\footnote{Verse 14. Pali: \textit{``Ekattatāya dhammā, yepi ca vemattatāya niddiṭṭhā; / Tena vikappayitabbā, eso hāro adhiṭṭhāno.''} \textit{Ekattatā} means ``sameness, identity of nature.'' \textit{Vemattatā} means ``difference, diversity of nature.'' \textit{Vikappayitabba} means ``should be distinguished, should be analyzed.'' \textit{Adhiṭṭhāna} comes from \textit{adhi} + \textit{ṭhāna} (standing upon) and here means ``determination, resolution'' --- deciding definitively whether two teachings point to the same thing or different things.}

For example: are ``calm'' (\textit{samatha}) and ``concentration'' (\textit{samādhi}) the same thing? Different words for one reality, or two distinct mental qualities? Your answer shapes your entire understanding of the path. The Determination Mode says: don't leave this question unresolved.

\subsection*{15. Requisites}

\begin{versebox}
Which qualities generate which quality,\\
as conditions in sequence ---\\
drawing out the cause: the Requisites Mode.
\end{versebox}

This mode traces the supply chain of the spiritual path. Which quality is the prerequisite for which? What do you need in order to get what you need?\footnote{Verse 15. Pali: \textit{``Ye dhammā yaṃ dhammaṃ, janayantippaccayā paramparato; / Hetumavakaḍḍhayitvā, eso hāro parikkhāro.''} \textit{Janayanti} (they generate), \textit{paccayā} (as conditions), \textit{paramparato} (in sequence, one after another). The key phrase is \textit{hetumavakaḍḍhayitvā} --- ``having drawn out the cause'' --- from \textit{hetu} (cause) + \textit{avakaḍḍhati} (to draw out, to pull). The image is of drawing a thread out of a tangle: you identify the causal chain and lay it out in order.}

This is enormously practical. Instead of seeing the path as a checklist (``develop these eight factors''), you see it as a causal sequence: this leads to this, which enables this, which produces this. Effort leads to mindfulness leads to concentration leads to wisdom.\footnote{The paradigmatic example is the \textit{upanisā sutta} (SN 12.23), which traces a chain from suffering through faith, joy, rapture, tranquility, happiness, concentration, and knowledge-and-vision of things as they really are, all the way to liberation. The Requisites Mode reads this chain and asks: what is the ``equipment'' (\textit{parikkhāra}) needed at each link?}

\subsection*{16. Integration}

\begin{versebox}
Teachings that share one root,\\
those the Sage proclaimed as having one aim ---\\
bring them together. That's the Integration Mode.
\end{versebox}

The final mode. After you've broken everything apart with modes 1 through 15, this mode puts it all back together. Teachings that share the same root, the same purpose, the same direction --- collect them. Synthesize.\footnote{Verse 16. Pali: \textit{``Ye dhammā yaṃ mūlā, ye cekatthā pakāsitā muninā; / Te samaropayitabbā, esa samāropano hāro.''} \textit{Ekatthā} means ``having one purpose, one meaning'' (\textit{eka} + \textit{attha}). \textit{Muni} (the Sage) is an epithet for the Buddha. \textit{Samaropayitabbā} means ``should be made to ascend together, should be integrated'' --- from \textit{sam} + \textit{ā} + \textit{ropeti} (to make climb). The image is of bringing dispersed items up to a single summit. This is the capstone mode: the move from analysis to synthesis, from dissection to integration.}

Analysis without synthesis is just a pile of parts. The Integration Mode is the moment you step back from the disassembled engine and put it back together --- only now you understand how every piece works.

\ornament

\section*{The Five Methods in Verse}

\subsection*{17. The Joyful Round}

\begin{versebox}
Craving and ignorance, led by calm and insight ---\\
yoked together with the Truths.\\
That's the Joyful Round.
\end{versebox}

The simplest method: take the master pair of problems (craving and ignorance) and the master pair of solutions (calm and insight), and connect them through the Four Noble Truths.\footnote{Verse 17. Pali: \textit{``Taṇhañca avijjampi ca, samathena vipassanā yo neti; / Saccehi yojayitvā, ayaṃ nayo nandiyāvaṭṭo.''} \textit{Neti} (leads, guides --- from the root \textit{nī}, to lead) is a play on the title \textit{Netti} itself. The method ``leads'' (\textit{neti}) craving (\textit{taṇhā}) and ignorance (\textit{avijjā}) by means of calm (\textit{samatha}) and insight (\textit{vipassanā}), ``having yoked them with the truths'' (\textit{saccehi yojayitvā}). The ``yoke'' metaphor (\textit{yoga}) connects to \textit{yoga} in its original sense: joining, harnessing.}

\subsection*{18. The Triple Tip}

\begin{versebox}
Leading the unwholesome with its roots,\\
and the wholesome with its roots ---\\
real, true, not otherwise: the Triple Tip.
\end{versebox}

Now the lens widens from two roots to six: greed, hatred, and delusion on the dark side; non-greed, non-hatred, and non-delusion on the bright side. Each pair is a ``tip'' of the trident.\footnote{Verse 18. Pali: \textit{``Yo akusale samūlehi, neti kusale ca kusalamūlehi; / Bhūtaṃ tathaṃ avitathaṃ, tipukkhalaṃ taṃ nayaṃ āhu.''} Three epithets describe this method's reliability: \textit{bhūta} (real, actual), \textit{tatha} (true, in accordance with reality --- the same root as \textit{Tathāgata}), and \textit{avitatha} (not false, not otherwise). These are strong epistemological claims: this method, the Netti says, shows things as they really are.}

Note those three words: ``real, true, not otherwise.'' The Netti is making a bold claim here --- this method isn't just useful, it's \textit{epistemologically valid}. It shows things as they actually are.

\subsection*{19. The Lion's Play}

\begin{versebox}
Leading the defilements through the distortions,\\
and the good qualities through the faculties ---\\
those who know methods call this the Lion's Play.
\end{versebox}

The distortions of perception --- seeing permanence where there's impermanence, pleasure where there's suffering, self where there's non-self, beauty where there's the unattractive --- these are the doorways into defilements. The faculties --- faith, energy, mindfulness, concentration, wisdom --- are the doorways into good qualities.\footnote{Verse 19. Pali: \textit{``Yo neti vipallāsehi, kilese indriyehi saddhamme; / Etaṃ nayaṃ nayavidū, sīhavikkīḷitaṃ āhu.''} \textit{Vipallāsa} (distortion, perversion of perception). \textit{Kilesa} (defilements). \textit{Indriya} (faculties). \textit{Saddhamma} (true Dhamma, good qualities). \textit{Nayavidū} (those who know the methods, the methodologically skilled). The lion (\textit{sīha}) surveys every direction before acting --- this method surveys the full terrain of defilements and their antidotes with the confidence of a lion in its domain.}

\subsection*{20. Scanning the Horizon}

\begin{versebox}
The wholesome and unwholesome spoken here and there\\
throughout the explanatory texts ---\\
surveying them mentally: that's Scanning the Horizon.
\end{versebox}

Before you analyze, survey. Scan the entire teaching and note every wholesome and unwholesome quality that appears, wherever it appears.\footnote{Verse 20. Pali: \textit{``Veyyākaraṇesu hi ye, kusalākusalā tahiṃ tahiṃ vuttā; / Manasā volokayate, taṃ khu disālocanaṃ āhu.''} \textit{Veyyākaraṇa} (explanation, exposition --- one of the nine divisions of the Buddha's teaching). \textit{Tahiṃ tahiṃ} (here and there, in various places). \textit{Manasā volokayate} (surveys mentally, looks with the mind). \textit{Disā} (direction, quarter) + \textit{ālocana} (examining, looking) --- looking in all directions, the way a sentinel scans the horizon.}

\subsection*{21. The Hook}

\begin{versebox}
Having surveyed with the Horizon Scan,\\
lifting out and bringing together\\
all the wholesome and unwholesome --- that's the Hook.
\end{versebox}

After scanning, you pull out what you've found and sort it. The Hook gathers, extracts, organizes.\footnote{Verse 21. Pali: \textit{``Oloketvā disalocanena, ukkhipiya yaṃ samāneti; / Sabbe kusalākusale, ayaṃ nayo aṅkuso nāma.''} \textit{Oloketvā} (having surveyed). \textit{Ukkhipiya} (having lifted up, having extracted). \textit{Samāneti} (brings together, gathers). \textit{Aṅkusa} is a goad or hook used by elephant trainers --- the mahout's tool for guiding an enormous animal with precision. The method hooks the essential elements out of the mass of text the way a mahout steers an elephant.}

\subsection*{22. How They All Fit Together}

\begin{versebox}
The sixteen modes come first;\\
then survey the directions with the Horizon Scan;\\
compress with the Hook;\\
and expound the sutta with the three methods.
\end{versebox}

This final verse is the operating manual for the entire system. First, apply the sixteen modes to break the text open. Then survey everything you've found. Then compress it with the Hook. Then explain what you've found using the Joyful Round, the Triple Tip, and the Lion's Play.\footnote{Verse 22. Pali: \textit{``Soḷasa hārā paṭhamaṃ, disalocanato disā viloketvā; / Saṅkhipiya aṅkusena hi, nayehi tīhi niddise suttaṃ.''} Note the sequence: (1) the sixteen hāras, (2) Scanning the Horizon, (3) the Hook, then (4) the three nayas (Joyful Round, Triple Tip, Lion's Play). Scanning the Horizon and the Hook are preparatory --- they survey and gather. The actual exposition (\textit{niddise suttaṃ}) happens through the three ``content'' methods. The variant reading \textit{disalocanena} (``by the Horizon Scan'') is attested in some manuscripts (\textit{ka.} = Cambodian edition).}

\ornament

\section*{The Twelve Building Blocks}

Now Mahākaccāyana shifts from \textit{tools} to \textit{materials}. What is a sutta actually \textit{made of}? He says: twelve elements, split into two groups of six.\footnote{This section is called \textit{Dvādasapada} --- ``The Twelve Terms'' (literally ``twelve-footed,'' since \textit{pada} means both ``term'' and ``foot'').}

\subsection*{The Six Elements of the Letter}

\begin{versebox}
Syllable, word, and phrasing,\\
grammar, description, and style ---\\
these six make up the entire letter.
\end{versebox}

These are the building blocks of the \textit{text itself} --- the surface layer, the actual words on the page (or in the ear).\footnote{Verse 23. Pali: \textit{``Akkharaṃ padaṃ byañjanaṃ, nirutti tatheva niddeso; / Ākārachaṭṭhavacanaṃ, ettāva byañjanaṃ sabbaṃ.''} The six terms: (1) \textit{akkhara} (syllable, letter --- the smallest unit of linguistic form), (2) \textit{pada} (word --- the basic meaningful unit), (3) \textit{byañjana} (phrasing, expression --- how words are combined), (4) \textit{nirutti} (grammatical analysis, etymology --- parsing the language), (5) \textit{niddesa} (description, exposition --- how the teaching is laid out), and (6) \textit{ākāra} (style, manner --- literally ``appearance'' or ``way of doing''). These move from the smallest linguistic unit (syllable) to the largest (style), forming a complete anatomy of the sutta's textual surface.}

\begin{enumerate}[itemsep=2pt]
\item \textbf{Syllable} --- the smallest unit. Individual sounds.
\item \textbf{Word} --- the basic meaningful unit.
\item \textbf{Phrasing} --- how words are combined into expressions.
\item \textbf{Grammar} --- how the language is parsed and analyzed.
\item \textbf{Description} --- how the teaching is laid out and explained.
\item \textbf{Style} --- the manner and form of expression.
\end{enumerate}

\subsection*{The Six Elements of the Meaning}

\begin{versebox}
Illustration, clarification, disclosure,\\
analysis, elucidation, and designation ---\\
by these six terms, the meaning is described.
\end{versebox}

These are the building blocks of what the text \textit{means} --- the deeper layer underneath the words.\footnote{Verse 24. Pali: \textit{``Saṅkāsanā pakāsanā, vivaraṇā vibhajanuttānīkammapaññatti; / Etehi chahi padehi, attho kammañca niddiṭṭhaṃ.''} The six terms: (1) \textit{saṅkāsanā} (illustration --- making something visible through comparison or example), (2) \textit{pakāsanā} (clarification --- making clear what was obscure), (3) \textit{vivaraṇā} (disclosure, opening up --- revealing what was hidden), (4) \textit{vibhajanā} (analysis --- breaking down into components), (5) \textit{uttānīkamma} (elucidation --- literally ``making face-up,'' i.e., making obvious what was face-down or concealed), and (6) \textit{paññatti} (designation --- naming, conceptualizing). These six move from concrete illustration to abstract designation, forming a complete anatomy of meaning-making.}

\begin{enumerate}[itemsep=2pt]
\item \textbf{Illustration} --- making a point through comparison or example.
\item \textbf{Clarification} --- making clear what was obscure.
\item \textbf{Disclosure} --- revealing what was hidden.
\item \textbf{Analysis} --- breaking something into its parts.
\item \textbf{Elucidation} --- turning the implicit into the obvious.
\item \textbf{Designation} --- naming and conceptualizing.
\end{enumerate}

\subsection*{The Grand Total}

\begin{versebox}
Three methods, plus six terms for meaning ---\\
nine terms for the meaning side.\\
Add twenty-four for textual investigation ---\\
thirty-three in all: that is the Guide.
\end{versebox}

Here Mahākaccāyana totals up his system.\footnote{Verses 25--26. Pali: \textit{``Tīṇi ca nayā anūnā, atthassa ca chappadāni gaṇitāni; / Navahi padehi bhagavato, vacanassattho samāyutto.''} --- ``Three methods, complete, plus six meaning-terms counted: by nine terms the meaning of the Blessed One's word is furnished.'' And: \textit{``Atthassa navappadāni, byañjanapariyeṭṭhiyā catubbīsa; / Ubhayaṃ saṅkalayitvā, tettiṃsā ettikā nettīti.''} --- ``Nine terms for meaning, twenty-four for textual investigation; combining both, thirty-three --- that is the extent of the Guide.'' The twenty-four for ``textual investigation'' (\textit{byañjanapariyeṭṭhi}) comprises the sixteen modes, the two remaining methods (Scanning the Horizon and the Hook), and the six terms for the letter (6+16+2=24). The variant reading \textit{saṅkhepayato} for \textit{saṅkalayitvā} (``summarizing'' for ``combining'') is attested in the Cambodian edition.} On the meaning side: three methods (Joyful Round, Triple Tip, Lion's Play) plus six meaning-terms equals nine. On the letter side: sixteen modes plus two methods (Scanning the Horizon and the Hook) plus six letter-terms equals twenty-four. Grand total: thirty-three.

That's the entire system. Thirty-three tools for reading any sutta the Buddha ever taught.

\bigskip

Notice the elegance of the architecture. The meaning side gives you \textit{what the Buddha is pointing at}. The letter side gives you \textit{how to read the words that do the pointing}. You need both --- and together they give you everything.

This is where the compressed overview ends and the deep dive begins. In the next chapter, we zoom in on the first mode --- Teaching --- and see exactly how it works on real suttas.
